Convective heat transport in stratified atmospheres at low and high Mach number

### 2017-03-09 tex/nusselt_paper.tex
gets all of the ideas into the discussion

We utilize the 
Dedalus\footnote{http://dedalus-project.org/} \cite{burns&all2016} 
pseudospectral framework to time-evolve  
(\ref{eqn:continuity_eqn})-(\ref{eqn:energy_eqn}) 
using an implicit-explicit (IMEX), third-order, four-step 
Runge-Kutta timestepping scheme RK443 \cite{ascher&all1997}.  
Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal, periodic) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  
Physical grid sizes range from
64x256 coefficient points at the lowest values of 
Ra to 1024x4096 coefficient points at Ra$_{\text{top}} > 10^{7}$ in 2D,
and 64x128$^2$ to  128x256$^2$ in 3D. 
By using IMEX timestepping, we implicitly step the 
stiff linear acoustic wave contribution and are able to
efficiently study flows at moderate ($\approx 1$) 
and very low ($\approx 10^{-4}$) Ma.  Our equations take the form
of the FC equations in \cite{lecoanet&all2014}, extended to include
$\nu$ and $\chi$ which vary with depth, and we follow the approach there.
This IMEX approach has been successfully 
tested against a nonlinear benchmark  of the compressible 
Kelvin-Helmholtz instability \cite{Lecoanet_et_al_2016_KH}.

@@

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/ma_v_Ra.png}
\caption{The mean adiabatic Mach number of long-time-averaged profiles
is shown.  Error bars show the full range of Ma over the depth of the
atmosphere.
(a) When $Ma < 1$ (below the black horizontal line), 
Ma $\propto \epsilon^{1/2}$.
t low ($10^2$) and moderate ($10^5$) values of Ra.
(b) When Ra is small, Ma $\propto Ra^{1/2}$ (dashed lines).
Once the mean Ma reaches 1 ($\epsilon = \{0.5, 1\}$), this scaling breaks down
and nearly all flows in the atmosphere are at Ma = 1.
At low $\epsilon$, a consistent power law of Ma $\propto Ra^{1/4}$ is retrieved.
\label{fig:ma_v_eps} }
\end{figure}

@@

Low Ma flows ($\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
While it has been suggested that pressure forces 
cause symmetry breaking in up- and downflows
\cite{hurlburt&all1984}, at low $\epsilon$ this 
effect seems to be secondary to flows obeying mass conservation as they traverse
the stratified medium.  
Our choice of fixed-temperature boundary conditions allows the flux at the 
boundaries to vary, and runs at $\text{Ra }> 10^4$
and $\epsilon = 10^{-4}$ exhibit long-term states of
flux disequilibrium.  
Roll solutions such as those pictured in Fig. \ref{fig:entropy_snapshots}a
are punctuated by states of vigorous shearing, similar to those previously
reported in two-dimensional RB convection \cite{goluskin&all2014}.  
During the roll states, 
the flux at the lower boundary exceeds that at the upper boundary
and energy is stored in the system.
During shearing states, 
convective transport is suppressed and Nu diminishes while
the flux at the upper boundary exceeds that at the lower boundary.  Long term
averages over both of these states produce flat flux profiles which
can be sensibly analyzed.

@@

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/re_and_nu_v_Ra.png}
\caption{
Variation of Nu as Ra increases at high and low $\epsilon$. 
At high $\epsilon$ (purple circles), 
a clear transition from the subsonic to supersonic regime 
is evident in the scaling
of Nu with Ra at $\text{Ra}_{\text{top}} \approx 10^4$.  
In the low $\epsilon$ regime (green diamonds and yellow squares), 
our observed Nu scalings collapse onto a similar line which is
indistinguishable from a $\text{Ra}_{\text{top}}^{2/7}$ scaling, 
as observed in RB convection \cite{johnston&doering2009}.  
Errors bars indicate the properly normalized range of the
value of Nu as a function of depth and
large error bars indicate a poorly converged solution.
 \label{fig:re_and_nu_v_ra}.
}
\end{figure}

@@

The variation of Nu with Ra$_{\text{top}}$ is shown in 
Fig. \ref{fig:re_and_nu_v_ra}a.
At low to moderate Ra$_{\text{top}}$, 
Nu $\propto$ Ra$_{\text{top}}^{1/3}$ regardless of $\epsilon$,
as predicted by classical \RB theory \cite{king&all2012}.
At large Ra$_{\text{top}}$, Nu $\propto$ Ra$_{\text{top}}^{1/5}$ 
at $\epsilon = 0.5$.
The scaling of Nu with Ra$_{\text{top}}$ 
is unclear at low $\epsilon$ (so I'll try to fill in that space with
data from comps...time to fish it out of Lou and do post-processing hell).

The Reynolds number and Peclet number,
\begin{equation}
\text{Re} = \frac{|\bm{u}| L_z}{\nu};\,\,\,\,\text{Pe} = \text{Pr}\,\text{Re},
\end{equation}
quantify the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in
Fig. \ref{fig:re_and_nu_v_ra}b.
At low Ra$_{\text{top}}$, Re $\propto$ Ra$_{\text{top}}^{3/4}$, but at high 
Ra$_{\text{top}}$ where the average Ma $\approx 1$ for $\epsilon = 0.5$, 
this scaling gives way to a Re $\propto$ Ra$_{\text{top}}^{1/2}$.

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/density_v_ra.png}
\caption{\label{fig:nrho_v_ra} We measure the stratification of 
evolved solutions
in two ways.  The solid symbols show $\ln(\rho(0)/\rho(L_z))$, 
or the density contrast
as measured at the upper and lower boundary.  The empty symbols show 
$\ln(\text{max}(\rho)/\text{min}(\rho))$. 
Unsurprisingly, at low $\epsilon$ the evolved
$n_{\rho}$ is not different from the initial conditions to first order.  
At high $\epsilon$,
the density contrast shrinks, and once the mean 
Ma approaches 1 the two methods of
measuring the density stratification bifurcate as density 
inversions form within the thermal
boundary layers.}
\end{figure}

@@

\section{Discussion}
\refstepcounter{section}
\label{sec:discussion}
In this letter we have studied fundamental heat transport by 
stratified convection in simplified 2-D polytropic atmospheres.
We argue that these atmospheres are the natural extension
of the RB problem to stratified systems, 
and are an ideal laboratory for understanding the basic 
properties of stratified convection. 
We see little difference between
the properties of our 2D and 3D results, which aligns with expectations
of Boussinesq theory and DNS results at values of Pr $\geq$ 1 \cite{ahlers&all2009}.

At low Ra and Ma, the scaling of Nu with Ra is reminiscent of RB convection.  
However, at high Ra and Ma, the scaling breaks down to a Ra$^{1/5}$ 
scaling unlike what is seen in Boussinesq convection.  
Similarly, the Reynolds number switches from a Ra$^{3/4}$ to
a Ra$^{1/2}$ scaling in this regime shift.  
It is interesting that both of these
scaling laws turn over by roughly a factor of 2/3.

Time-dependent oscillating shear states have developed spontaneously, 
as seen before in RB convection \cite{goluskin&all2014}.  
While computationally difficult, 
the highest values of Ra and the lowest value
of $\epsilon$ studied here are far from values found in nature.  
Future work will aim to better understand the mechanisms of shearing states and
whether or not these states are attainable in 
three-dimensional, non-rotating atmospheres.

Furthermore, we have found that the stratification of 
these atmospheres evolves in a complex
manner, and future work should aim to 
understand the importance of changing stratification on
convective heat transport. Our studies
here will serve as a foundation both for understanding and 
comparing heat transport in stratified convection
to that in RB convection \cite{johnston&doering2009}, 
and for future studies of transport in stratified convection.  
These results can be used to determine if simplified equation sets, 
such as the anelastic equations, carry heat in the same manner as the 
FC equations.
This work will also be useful in coming to understand more realistic systems, 
such as rapidly rotating atmospheres \cite{julien&all2012},
atmospheres bounded by stable regions \cite{hurlburt&all1986}, 
or regions with realistic profiles of $\kappa$.

>>>>

We utilize the 
Dedalus\footnote{http://dedalus-project.org/} \cite{burns&all2016} 
pseudospectral framework to time-evolve  
(\ref{eqn:continuity_eqn})-(\ref{eqn:energy_eqn}) 
using an implicit-explicit (IMEX), third-order, four-step 
Runge-Kutta timestepping scheme RK443 \cite{ascher&all1997}.  
Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal, periodic) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  
Physical grid sizes range from
64x256 coefficient points at the lowest values of 
Ra to 1024x4096 coefficient points at Ra$_{\text{top}} > 10^{7}$ in 2D,
and 64x128$^2$ to  256x512$^2$ in 3D. 
By using IMEX timestepping, we implicitly step the 
stiff linear acoustic wave contribution and are able to
efficiently study flows at moderate ($\approx 1$) 
and very low ($\approx 10^{-4}$) Ma.  Our equations take the form
of the FC equations in \cite{lecoanet&all2014}, extended to include
$\nu$ and $\chi$ which vary with depth, and we follow the approach there.
This IMEX approach has been successfully 
tested against a nonlinear benchmark  of the compressible 
Kelvin-Helmholtz instability \cite{Lecoanet_et_al_2016_KH}.

@@

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/ma_v_Ra.png}
\caption{The mean adiabatic Mach number of long-time-averaged profiles
is shown.  Error bars show the full range of Ma over the depth of the
atmosphere.
(a) Ma is plotted as a function of $\epsilon$, the superadiabatic excess,
at various values of Ra.
(b) Ma is plotted as a function of Ra at various values of $\epsilon$.
Larger symbols with white dots designate 3D runs.
\label{fig:ma_v_eps} }
\end{figure}

@@

Low Ma flows ($\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
While it has been suggested that pressure forces 
cause symmetry breaking in up- and downflows
\cite{hurlburt&all1984}, at low $\epsilon$ this 
effect seems to be secondary to flows obeying mass conservation as they traverse
the stratified medium.  
Our choice of fixed-temperature, stress free boundary conditions 
allows the flux at the boundaries to vary, while also allowing
for large-scale ``axisymmetric'' horizontal velocities.
Runs at $\text{Ra }> 10^4$
and $\epsilon = 10^{-4}$ exhibit long-term states of
flux disequilibrium, in which roll solutions such as those pictured in 
Fig. \ref{fig:entropy_snapshots}a
are punctuated by shearing states similar to those previously
reported in two-dimensional RB convection \cite{goluskin&all2014}.  
During both of these states, the system is in flux disequilibrium.
A long time average
over both of these states produce flat flux profiles which
can be sensibly analyzed, but the presence of these time-dependent states
makes it difficult to measure fluid quantities in these cases.

@@

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/re_and_nu_v_Ra.png}
\caption{
(a) Variation of Nu as Ra increases at high and low $\epsilon$. 
Errors bars indicate the properly normalized range of the
value of Nu as a function of depth and
large error bars indicate a poorly converged solution.
(b) Variation of Re as Ra increases at high and low $\epsilon$.
Re is measured at the midplane of the atmosphere.
Larger symbols with white dots designate 3D runs.
 \label{fig:re_and_nu_v_ra}.
}
\end{figure}

@@

The variation of Nu with Ra$_{\text{top}}$ is shown in 
Fig. \ref{fig:re_and_nu_v_ra}a.
At low to moderate Ra$_{\text{top}}$, 
Nu $\propto$ Ra$_{\text{top}}^{1/3}$ regardless of $\epsilon$,
as predicted by classical \RB theory \cite{king&all2012}.
At large Ra$_{\text{top}}$, Nu $\propto$ Ra$_{\text{top}}^{1/5}$ 
at $\epsilon = 0.5$.
The scaling of Nu with Ra$_{\text{top}}$ 
is unclear at low $\epsilon$ once Ra is sufficiently large
for shearing states to occur.  It would be beneficial to study
low $\epsilon$ flows with fixed-flux boundary conditions or
no-slip boundary conditions that disallow these large scale
structures.

The Reynolds number and Peclet number,
\begin{equation}
\text{Re} = \frac{|\bm{u}| L_z}{\nu};\,\,\,\,\text{Pe} = \text{Pr}\,\text{Re},
\end{equation}
quantify the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in
Fig. \ref{fig:re_and_nu_v_ra}b.
At low Ra$_{\text{top}}$, Re $\propto$ Ra$_{\text{top}}^{3/4}$, but at high 
Ra$_{\text{top}}$ where the average Ma $\approx 1$ for $\epsilon = 0.5$, 
this scaling gives way to a Re $\propto$ Ra$_{\text{top}}^{1/2}$.
For large values of Ra but at low $\epsilon$, there appears to be a scaling
of Re $\propto$ Ra$^{2/3}$.

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/density_v_ra.png}
\caption{\label{fig:nrho_v_ra} The stratification of 
evolved solutions is measured
in two ways.  Solid symbols show $\ln(\rho(z=0)/\rho(z=L_z))$, 
the density contrast
as measured at the upper and lower boundary.  The empty symbols show 
$\ln(\text{max}(\rho)/\text{min}(\rho))$. 
Unsurprisingly, at low $\epsilon$ the evolved
$n_{\rho}$ is not different from the initial conditions to first order.  
At high $\epsilon$,
the density contrast shrinks, and once the mean 
Ma approaches 1 (as in Fig. \ref{fig:ma_v_eps}b), the two methods of
measuring the density stratification bifurcate as density 
inversions form within the thermal
boundary layers. Larger symbols with white dots represent 3D runs.}
\end{figure}

@@

\section{Discussion}
\refstepcounter{section}
\label{sec:discussion}
In this letter we have studied fundamental heat transport by 
stratified convection in simplified 2D and 3D polytropic atmospheres.
We argue that these atmospheres are the natural extension
of the RB problem to stratified systems, 
and are an ideal laboratory for understanding the basic 
properties of stratified convection. 
We see little difference between
the properties of our 2D and 3D results, which aligns with expectations
of Boussinesq theory and DNS results at values of Pr $\geq$ 1 \cite{ahlers&all2009}.

At low Ra and Ma, the scaling of Nu with Ra is reminiscent of RB convection.  
However, at high Ra and Ma, the scaling of Nu becomes weakear, decreasing
from a 1/3 to a 1/5 power law, a decrease by roughly a factor of 1/3.
Similarly, the Reynolds number switches from a 3/4 power law to
a 1/2 scaling in this regime shift, a similar decrease by a factor of 1/3.
This seems to suggest that a large portion of the scaling of the Nusselt number
comes from increases in velocity as Ra increases.  Once the velocity reaches
its natural maximum, heat transport can no longer benefit from increased velocities.
The same is true of the Reynolds number, which tracks increases in velocities as
the flows become more turbulent, but which can only scale with the diffusivities
once the Mach number reaches its maximum value.

Time-dependent oscillating shear states have developed spontaneously, 
as seen before in RB convection \cite{goluskin&all2014}, and these should
be studied in more detail.  These states are fundamentally in flux disequilibrium
and make it very difficult to calculate a meaningful Nusselt number.  It would
be beneficial for future studies to study solutions with fixed flux or no slip
boundary conditions to better converge and study cases at high $\epsilon$ and low
Ra.

Finally, we have found that the stratification of 
these atmospheres evolves in a complex
manner.  Future work should aim to 
understand the importance of stratification on
convective heat transport, and the manner in which
boundary layer density inversions interact with
thermal plumes.

Our studies
here will serve as a foundation both for understanding and 
comparing heat transport in stratified convection
to that in RB convection \cite{johnston&doering2009}, 
and for future studies of transport in stratified convection.  
These results can be used to determine if simplified equation sets, 
such as the anelastic equations, carry heat in the same manner as the 
FC equations.
This work will also be useful in coming to understand more realistic systems, 
such as rapidly rotating atmospheres \cite{julien&all2012},
atmospheres bounded by stable regions \cite{hurlburt&all1986}, 
or regions with realistic profiles of $\kappa$.
